Water, High-Altitude Condensates, and Methane Depletion in the Atmosphere of the Warm Neptune WASP-107\lowercase{b}

### 2017-08-14 ms.tex
draft 1

\begin{abstract}
The transiting planet WASP-107b is an intriguing target for atmosphere characterization, with a large atmospheric scale height and a small, bright host star.

w107 is a good target for atmospheric characterization, raising the exciting possibility of characterizing its formation history

.  With a mass similar to Neptune's and a radius close to Jupiter's, it occupies the transition region between ice giants and gas giants. The planet is also one of the strongest targets for atmospheric characterization yet discovered, thanks to its small, bright host star and large scale height.  We report the first atmosphere characterization of WASP-107b, a Hubble Space Telescope measurement of its near-infrared transmission spectrum.  We analyzed the planet's composition with two techniques: atmospheric retrieval based on the transmission spectrum and interior structure modeling based on the planet's mass and radius. We detect water absorption at $6.5\,\sigma$ confidence and set an upper limit on the atmospheric metallicity of $30\times$ solar. We find that the methane abundance is depleted relative to expectations for a solar abundance pattern (at $3\,\sigma$ confidence), suggesting a low carbon-to-oxygen ratio or high internal heat flux.  The amplitude of features in the transmission spectrum is smaller than expected for a cloudless atmosphere, crossing just one scale height, and requires an opaque condensate layer at high altitudes (0.1 - 3 mbar). We find that it is challenging for physically motivated cloud and haze models to produce opaque condensates at this height.  Taken together, these findings serve as an illustration of the diversity and complexity of exoplanet atmospheres. We can look forward to more such results with the high precision and wide spectral coverage afforded by future observing facilities. 
\end{abstract}

@@

WASP-107b also has a relatively low equilibrium temperature (780 K) compared to most other exoplanets that are amenable to atmosphere characterization.  This results in a distinct atmospheric chemistry from other well-studied systems: at low temperatures, the dominant molecular reservoir for carbon transitions from carbon monoxide to methane \citep{moses13}.  Spectral features from both water and methane are accessible with current observing facilities, raising the possibility of a spectroscopic estimate of the carbon-to-oxygen ratio (C/O). Previous measurements of C/O have been challenging because they rely on photometry that covers absorption features from multiple molecules \citep[e.g.][]{madhusudhan11, line14, benneke15, kreidberg15b}. FIXME: maybe cut this.

@@

\section{Observations}
We observed a single transit of WASP-107b with HST's Wide Field Camera 3 (WFC3) instrument on UT 5-6 June 2017.  The transit observation consisted of five HST orbits. At the beginning of each 96-minute orbit, we took an image of the target with the F130N filter (exposure time = 4.2 s). This direct image is used for wavelength calibration. For the remainder of the target visibility period (about 45 minutes), we obtained time series spectra with the G141 grism, which provides low-resolution spectroscopy over the wavelength range $1.1 - 1.7\,\mu$m.  We used the NSAMP=6, SPARS\_25 readout mode (exposure time = 112 s) to optimize the efficiency of the observations, as determined by the \texttt{PandExo\_HST} planning tool\footnote{https://github.com/spacetelescope/PandExo\_HST}.  As is standard for WFC3 observations of bright targets, we used the spatial scanning observing mode, which slews the telescope in the spatial direction over the course of an exposure. The scan rate was 0.12 arcseconds/sec.

@@

We explored several alternative choices for the spectroscopic light curve fits, but found that none of them made a significant difference in the transmission spectrum. For example, we fit the spectroscopic light curves with the same analytic model we used for the white light curve and obtained nearly identical relative transit depths (differing by just $0.3\sigma$ on average), except with a small constant offset due to the uncorrected star-spot crossing feature. This offset does not affect our final analysis because the planet's one-bar radius is a free parameter in the atmospheric retrieval (see \S\,\ref{sec:retrieval}). We also fit for a linear limb darkening parameter rather than fixing the limb darkening on the PHOENIX model values. We found that the fitted limb darkening coefficients are consistent with the model predictions, so we opted to fix the coefficients in our final analysis because it improves the precision on the transit depths by about 10\%. We also checked that the uncertainty in the stellar parameters does not significantly affect the PHOENIX model predictions at the level of precision of our data.

@@

\subsubsection{Independent Analysis}
We also performed an independent data reduction and fit using K. Stevenson's pipeline.

@@

\begin{figure}
\includegraphics[width = 0.5\textwidth]{Figures/metalByMass.png}
\caption{Estimated envelope metallicity and core mass based on interior structure modeling of the mass and radius of WASP-107b. FIXME ages.}
\label{fig:metal_prior}
\end{figure}

@@

The nominal retrieval results are shown in Figure\,\ref{fig:retrieval}.  The metallicity distribution spans the full range allowed by our priors, with some preference for larger values. The cloud top pressure is estimated to be $0.01 - 3$ mbar at $1\,\sigma$ confidence. These are not particularly precise constraints because the metallicity and cloud height are somewhat degenerate \citep{griffith14}.  The retrieved irradiation temperature is $525 - 820$ K (1 $\sigma$ confidence).  We find that the C/O value is less than solar (0.54) at $2.7\,\sigma$ confidence.

@@

\subsection{Condensate Properties}
In addition to the atmospheric retrieval, we also performed forward modeling of physically motivated, self-consistent clouds and hazes using the methods described in \cite{fortney08, morley15}.  We model clouds that form in cool atmospheres (Na$_2$S, KCl, ZnS; see \citealt{morley12}), for a range of metallicites from $1-50\times$ solar. We vary the cloud sedimentation efficiency from 0.1 to 3 (highly lofted, small particulate clouds). None of the models produce sufficiently low amplitude spectral features to match the observed spectrum. FIXME: what did we assume about surface gravity?

@@

\begin{itemize}
\item{\emph{The upper limit on atmospheric metallicity is $30\times$ solar.} Our estimated upper limit is at the edge of consistency with the Solar System metal enrichment pattern, with predicts a metallicity of $30\times$ solar for WASP-107b \citep{kreidberg14b}.  If the star is old (FIXME Gyr), the upper limit on metallicity would be even lower, implying that WASP-107b has a different formation history from the Solar System planets. Further characterization of the stellar age is needed to explore this scenario.  Either way, this planet stands in contrast to the similar mass and temperature planet GJ 436b, which is thought to have a metallicity greater than $100\times$ solar \citep{morley17}. Together, these exoplanets hint at a greater diversity of compositions than exists in the Solar System.}
\item{The methane abundance is depleted relative to expectations for a solar abundance pattern. This may be due to an instrinsically low carbon-to-oxygen ratio, which could arise from accretion of water-rich planetesimals \citep{mordasini16, espinoza17}.  Another possibility is that the planet has a hot interior ($\sim500$ K), and abundances are quenched at pressure levels where CO is stable \citep[as observed in some directly imaged planets;][]{skemer14, zahnle14}. Such a high internal temperature could be due to latent heat of formation if the planet is very young, and/or tidal heating \citep{fortney08, morley17}. Further observations of the transmission spectrum over a broader wavelength range will refine the C/O estimate and help differentiate between these two scenarios.} 
\item{Optically thick condensates are present at high altitudes ($0.1 - 3$ mbar).  Existing models of physically-motivated cloud and haze formation do not satisfactorily reproduce the observations. The atmospheric scale height is so large (800 km) that clouds form too deep in the atmosphere to loft to the needed height. Hydrocarbon hazes that form at high altitudes can match the data, but these require methane photolysis as a precursor, and methane is depleted in the atmosphere.  Other haze precursors such as $H_2S$ are a possibility and would be worth exploring in future analyses \citep{gao17}.
\citep{crossfield17}} GJ 1214b
\end{itemize}

WASP-107b is slated to be observed with numerous facilities. Transits were recently observed with the WFC3/G102 grism and Spitzer 3.6 and 4.5 $\mu$m channels. Spitzer eclipses are planned (Program FIXME; PI L. Kreidberg). In addition, WASP-107b is included in the \emph{JWST} Guaranteed Time Observations.

@@

\vspace{5mm}
\facilities{HST(WFC3)}

%% Similar to \facility{}, there is the optional \software command to allow 
%% authors a place to specify which programs were used during the creation of 
%% the manusscript. Authors should list each code and include either a
%% citation or url to the code inside ()s when available.

%\software{astropy \citep{2013A&A...558A..33A},  
%          Cloudy \citep{2013RMxAA..49..137F}, 
%          SExtractor \citep{1996A&AS..117..393B}
%          }

%% Appendix material should be preceded with a single \appendix command.
%% There should be a \section command for each appendix. Mark appendix
%% subsections with the same markup you use in the main body of the paper.

%% Each Appendix (indicated with \section) will be lettered A, B, C, etc.
%% The equation counter will reset when it encounters the \appendix
%% command and will number appendix equations (A1), (A2), etc. The
%% Figure and Table counter will not reset.

%\appendix

%% The reference list follows the main body and any appendices.
%% Use LaTeX's thebibliography environment to mark up your reference list.
%% Note \begin{thebibliography} is followed by an empty set of
%% curly braces.  If you forget this, LaTeX will generate the error
%% "Perhaps a missing \item?".
%%
%% thebibliography produces citations in the text using \bibitem-\cite
%% cross-referencing. Each reference is preceded by a
%% \bibitem command that defines in curly braces the KEY that corresponds
%% to the KEY in the \cite commands (see the first section above).
%% Make sure that you provide a unique KEY for every \bibitem or else the
%% paper will not LaTeX. The square brackets should contain
%% the citation text that LaTeX will insert in
%% place of the \cite commands.

%% We have used macros to produce journal name abbreviations.
%% \aastex provides a number of these for the more frequently-cited journals.
%% See the Author Guide for a list of them.

%% Note that the style of the \bibitem labels (in []) is slightly
%% different from previous examples.  The natbib system solves a host
%% of citation expression problems, but it is necessary to clearly
%% delimit the year from the author name used in the citation.
%% See the natbib documentation for more details and options.

@@

%% This command is needed to show the entire author+affilation list when
%% the collaboration and author truncation commands are used.  It has to
%% go at the end of the manuscript.
%\allauthors

%% Include this line if you are using the \added, \replaced, \deleted
%% commands to see a summary list of all changes at the end of the article.
%\listofchanges

@@

% End of file `sample61.tex'.

>>>>

\begin{abstract}
The Neptune-mass exoplanet WASP-107b is an exciting target for atmosphere characterization. It has an unusually large atmospheric scale height and a small, bright host star, raising the possibility of precise constraints on its current nature and formation history.  We report the first atmospheric study of WASP-107b, a Hubble Space Telescope measurement of its near-infrared transmission spectrum.  We analyzed the planet's composition with two techniques: atmospheric retrieval based on the transmission spectrum and interior structure modeling based on the observed mass and radius. We detect water absorption at $6.5\,\sigma$ confidence and set an upper limit on the atmospheric metallicity of $30\times$ solar. We find that the methane abundance is depleted relative to expectations for a solar abundance pattern (at $3\,\sigma$ confidence), suggesting a low carbon-to-oxygen ratio or high internal heat flux.  The amplitude of features in the transmission spectrum is smaller than expected for a cloudless atmosphere, crossing less than one scale height, and requires a thick condensate layer at high altitudes (0.1 - 3 mbar). We find that it is challenging for physically motivated cloud and haze models to produce opaque condensates at this height.  Taken together, these findings serve as an illustration of the diversity and complexity of exoplanet atmospheres. We can look forward to more such results with the high precision and wide spectral coverage afforded by future observing facilities. 
\end{abstract}

@@

WASP-107b also has a relatively low equilibrium temperature (780 K) compared to most other exoplanets that are amenable to atmosphere characterization.  This results in a distinct atmospheric chemistry from other well-studied systems: at low temperatures, the dominant molecular reservoir for carbon transitions from carbon monoxide to methane \citep{moses13}.  Spectral features from both water and methane are accessible with current observing facilities, enabling a spectroscopic estimate of the carbon-to-oxygen ratio (C/O). Previous measurements of C/O have been challenging because they rely on photometry that covers absorption features from multiple molecules \citep[e.g.][]{madhusudhan11, line14, benneke15, kreidberg15b}.

@@

\section{Observations}
We observed a single transit of WASP-107b with HST's Wide Field Camera 3 (WFC3) instrument on UT 5-6 June 2017.  The transit observation consisted of five HST orbits. At the beginning of each 96-minute orbit, we took an image of the target with the F130N filter (exposure time = 4.2 s). This direct image is used for wavelength calibration. For the remainder of the target visibility period (about 45 minutes), we obtained time series spectra with the G141 grism, which provides low-resolution spectroscopy over the wavelength range $1.1 - 1.7\,\mu$m.  We used the NSAMP=6, SPARS\_25 readout mode (exposure time = 112 s) to optimize the efficiency of the observations, as determined by the \texttt{PandExo\_HST} planning tool\footnote{\url{https://github.com/spacetelescope/PandExo\_HST}}.  As is standard for WFC3 observations of bright targets, we used the spatial scanning observing mode, which slews the telescope in the spatial direction over the course of an exposure. The scan rate was 0.12 arcseconds/sec.

@@

We explored several alternative choices for the spectroscopic light curve fits, but found that none of them made a significant difference in the transmission spectrum. For example, we fit the spectroscopic light curves with the same analytic model we used for the white light curve and obtained nearly identical relative transit depths (differing by just $0.3\sigma$ on average), except with a small constant offset due to the uncorrected star-spot crossing feature. This offset does not affect our final analysis because the planet's one-bar radius is a free parameter in the atmospheric retrieval (see \S\,\ref{sec:retrieval}). We also fit for a linear limb darkening parameter rather than fixing the limb darkening on the PHOENIX model values. We found that the fitted limb darkening coefficients are consistent with the model predictions, so we opted to fix the coefficients in our final analysis because it improves the precision on the transit depths by about 10\%. We also checked that the uncertainty in the stellar parameters does not significantly affect the PHOENIX model predictions at the level of precision of our data.  We also performed an independent data reduction and fit using K. Stevenson's pipeline and found consistent results (well within $1\,\sigma$).

@@

@@

\begin{figure}
\includegraphics[width = 0.5\textwidth]{Figures/metalByMass.png}
\caption{Estimated envelope metallicity and core mass based on interior structure modeling of the mass and radius of WASP-107b. FIXME get new figure showing two ages.}
\label{fig:metal_prior}
\end{figure}

@@

The nominal retrieval results are shown in Figure\,\ref{fig:retrieval}.  The metallicity distribution spans the full range allowed by our priors, with some preference for larger values. The cloud top pressure is estimated to be $0.01 - 3$ mbar at $1\,\sigma$ confidence. These are not particularly precise constraints because the metallicity and cloud height are somewhat degenerate \citep{griffith14}.  The retrieved irradiation temperature is $525 - 820$ K (1 $\sigma$ confidence).  We find that the C/O value is less than solar (0.54) at $2.7\,\sigma$ confidence.  We tested the detection significance for water by removing water opacity from the nominal model. The Bayesian evidence favors the inclusion of water at $6.5\,\sigma$ confidence.

@@

\subsection{Condensate Properties}
In addition to the atmospheric retrieval, we also performed forward modeling of physically motivated, self-consistent clouds and hazes using the methods described in \cite{fortney08, morley15}.  We model clouds that form in cool atmospheres (Na$_2$S, KCl, ZnS; see \citealt{morley12}), for a range of metallicites from $1-50\times$ solar. We vary the cloud sedimentation efficiency from 0.1 to 3 (highly lofted, small particulate clouds). None of the models produce sufficiently low amplitude spectral features to match the observed spectrum. The results are sensitive to surface gravity, but even when we increase the gravity by FIXME $\sigma$ the clouds still form too low.

@@

\begin{itemize}
\item{\emph{The upper limit on atmospheric metallicity is $30\times$ solar.} This limit is at the edge of consistency with the Solar System metal enrichment pattern, which predicts a metallicity of $30\times$ solar for WASP-107b \citep{kreidberg14b}.  If the star is old ($>1$ Gyr), the upper limit on metallicity would be even lower (FIXME), implying that WASP-107b has a different formation history from the Solar System planets. Further characterization of the stellar age is needed to explore this scenario.  Either way, this planet stands in contrast to the similar mass and temperature planet GJ 436b, which is thought to have a metallicity greater than $100\times$ solar \citep{morley17}. The difference between WASP-107b and GJ 436b suggests a greater diversity of compositions for exoplanets than is present in the Solar System.}
\item{\emph{The methane abundance is depleted relative to expectations for a solar abundance pattern, whereas water is consistent with solar composition.} This may be due to an instrinsically low carbon-to-oxygen ratio, which could arise from accretion of water-rich planetesimals \citep{mordasini16, espinoza17}.  Another possibility is that the planet has a hot interior ($\sim500$ K), and abundances are quenched at pressure levels where CO is stable \citep[as observed in some directly imaged planets;][]{skemer14, zahnle14}. Such a high internal temperature could be due to latent heat of formation if the planet is very young, and/or tidal heating \citep{fortney08, morley17}. Further observations of the transmission spectrum over a broader wavelength range will refine the C/O estimate and help differentiate between these two scenarios.} 
\item{\emph{Optically thick condensates are present at high altitudes} ($0.1 - 3$ mbar). Existing models of physically-motivated cloud and haze formation do not satisfactorily reproduce the observed transmission spectrum. The atmospheric scale height is so large (800 km) that clouds form too deep in the atmosphere to loft to the needed height. Hydrocarbon hazes that form at high altitudes can match the data, but these require methane photolysis as a precursor, and we found that methane is depleted.  Other haze precursors such as H$_2$S are a possibility and would be worth exploring in future analyses \citep{gao17}. Put in context with other systems, the muted water feature for WASP-107b agrees well with the trend in feature amplitude with temperature noted in \cite{crossfield17}, indicating that condensates may be common in the coolest planets.}
\end{itemize}

These results are a first look at the atmosphere of WASP-107b. The planet is already being observed at other wavelengths, including the WFC3/G102 grism and Spitzer 3.6 and 4.5 $\mu$m channels in transit and eclipse (HST Program GO 14916, PI J. Spake, Spitzer Program 13052, PI M. Werner; Spitzer Program 13167, PI L. Kreidberg).  In addition, WASP-107b is included in the \emph{JWST} Guaranteed Time Observations for the NIRISS and NIRSpec instruments\footnote{\url{https://jwst-docs.stsci.edu}}.  This spate of observing programs is sure to add to the already rich and complex picture of WASP-107b's atmosphere presented here.

@@

@@

@@
